Finiteness and Cardinality in Homotopy Type Theory

### 2020-01-09 paper.tex
initial agda
now totally moved over

Finally, we can define formally split enumerability.
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}
\subsection{Split Surjections}
\begin{rm-definition}[Surjections]
  \begin{multicols}{2}
    \begin{equation}
      A \twoheadrightarrow B = \Sigma (f : A \rightarrow B) , \text{surjective}(f)
    \end{equation} \break
    \begin{equation}
      A \twoheadrightarrow! \; B = \Sigma (f : A \rightarrow B) , \text{split surjective}(f)
    \end{equation}
  \end{multicols}
  Where the definitions of surjective are taken from \cite[definition
  4.6.1]{hottbook}.
\end{rm-definition}
\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\end{proof}

\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, we know that
this implies split enumerable types must all be sets.
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
In order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma} \label{split-enum-bool-top-bot}
  \(\mathbf{Bool}\), \(\top\), and \(\bot\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  These non-recursive types have similar, simple proofs.
  \begin{multicols}{3}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{unit}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{empty}
  \end{multicols}
\end{proof}
\begin{rm-theorem} \label{split-enum-sigma}
  Split-enumerability is closed under \(\Sigma\).
  \begin{equation} \label{split-enum-sigma-clos-sig}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi (x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma (x : A) , U(x))
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-enum}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
Closure under disjoint union and Cartesian product both follow from \(\Sigma\),
as these types can be defined in terms of \(\Sigma\).
\begin{align}
  A \times B &= \Sigma {(x : A)} , B \\
  A + B &= \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
\end{align}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
Where split enumerability was the enumeration form of a split surjection from
\(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
\emph{equivalence} with \(\mathbf{Fin}\).

@@

  Kuratowski finite sets are closed under \(\times\) and \(+\).
  These follow from their lifted versions on manifestly enumerable types.
\todo[inline]{Fill in rest}

@@

For proof search, the procedure is a well-known one in Agda
\cite{devrieseBrightSideType2011}: we ask for the result of a decision procedure
as an \emph{instance argument}, which will demand computation during
typechecking.
As an example, we can check that the group operation on \(\AgdaDatatype{Pauli}\)
is idempotent like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{noninst-prf}

As a quick aside, when the property is \emph{not} true (for instance \(\forall x
\rightarrow x \cdot x \equiv x\)), Agda will fail by not finding an instance of
\(\bot\).
The error message specifically is
\verb+No instance of type ⊥ was found in scope+, which isn't very helpful.
We can do a little better by displaying the particular element on which the
property failed, by defining a custom empty type parameterised by the
counterexample itself.
Combined with some other changes to the combinators, this will give the
following more helpful error message:
\verb+No instance of type Counterexample X was found in scope+.
\subsection{Multiple Arguments}
The automation machinery above only deals with single-argument predicates.
This is not a problem, as we know that we can work with multiple arguments by
currying and uncurrying, since all of the finiteness predicates are closed under
\(\times\).
To automate away the curry/uncurry noise we will use instance search, building
on \cite{allaisGenericLevelPolymorphic2019} to develop a small interface to
generic \(n\)-ary functions and properties.
Our generic representation can handle dependent \(\Sigma\) and \(\prod\)
types (rather than their non-dependent counterparts, \(\times\) and
\(\rightarrow\)).
This extension was necessary for our use case: it is mentioned in the paper
as the obvious next step. 
We also implement the curry-uncurry combinators as (verified) isomorphisms.

A full explanation of our implementation is beyond the scope of this work, so we
only present the finished interface, which is used like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{prf-curried}

>>>>

Finally, we can define formally split enumerability.
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}
\subsection{Split Surjections}
\begin{rm-definition}[Surjections]
  \begin{multicols}{2}
    \begin{equation}
      A \twoheadrightarrow B = \Sigma (f : A \rightarrow B) , \text{surjective}(f)
    \end{equation} \break
    \begin{equation}
      A \twoheadrightarrow! \; B = \Sigma (f : A \rightarrow B) , \text{split surjective}(f)
    \end{equation}
  \end{multicols}
  Where the definitions of surjective are taken from \cite[definition
  4.6.1]{hottbook}.
\end{rm-definition}
\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \iff \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\end{proof}

\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, we know that
this implies split enumerable types must all be sets.
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
In order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma} \label{split-enum-bool-top-bot}
  \(\mathbf{Bool}\), \(\top\), and \(\bot\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  These non-recursive types have similar, simple proofs.
  \begin{multicols}{3}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{top}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bot}
  \end{multicols}
\end{proof}
\begin{rm-theorem} \label{split-enum-sigma}
  Split-enumerability is closed under \(\Sigma\).
  \begin{equation} \label{split-enum-sigma-clos-sig}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi (x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma (x : A) , U(x))
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-sup}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
Closure under disjoint union and Cartesian product both follow from \(\Sigma\),
as these types can be defined in terms of \(\Sigma\).
\begin{align}
  A \times B &= \Sigma {(x : A)} , B \\
  A + B &= \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
\end{align}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
Where split enumerability was the enumeration form of a split surjection from
\(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
\emph{equivalence} with \(\mathbf{Fin}\).

@@

We already have most of the components we need: closure under \(\bot\),
\(\top\), \(\mathbf{Bool}\), \(+\), \(\times\), and \(\rightarrow\).
What remains is the subobject classifier.
\begin{definition}
\end{definition}

@@

For proof search, the procedure is a well-known one in Agda
\cite{devrieseBrightSideType2011}: we ask for the result of a decision procedure
as an \emph{instance argument}, which will demand computation during
typechecking.
\subsection{Multiple Arguments}
The automation machinery above only deals with single-argument predicates.
This is not a problem, as we know that we can work with multiple arguments by
currying and uncurrying, since all of the finiteness predicates are closed under
\(\times\).
To automate away the curry/uncurry noise we will use instance search, building
on \cite{allaisGenericLevelPolymorphic2019} to develop a small interface to
generic \(n\)-ary functions and properties.
Our generic representation can handle dependent \(\Sigma\) and \(\prod\)
types (rather than their non-dependent counterparts, \(\times\) and
\(\rightarrow\)).
This extension was necessary for our use case: it is mentioned in the paper
as the obvious next step. 
We also implement the curry-uncurry combinators as (verified) isomorphisms.

A full explanation of our implementation is beyond the scope of this work, so we
only present the finished interface, which is used like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{assoc-prf}
